---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---

##Introduction

Course: Data Science Specification, Level: Reproducible Research.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. More information at:
https://github.com/monika66/RepData_PeerAssessment1/blob/master/README.md

The  activity monitoring data "activity.zip" for this report are cloned from forked GitHub repository: https://github.com/monika66/RepData_PeerAssessment1/blob/master/activity.zip 
(Original https://github.com/rdpeng/RepData_PeerAssessment1, state of 11 Feb 2014)


## Loading and preprocessing the data 

This data analysis was executed with RStudio, R version 3.3.3 (2017-03-06)

Windows platform: x86_64-w64-mingw32/x64 (64-bit)

CPU: Intel(R) Core(TM) i7-7500U, RAM: 16 GB 

For reproduction set environment workspace to your own workdirectory !
```{r environment, echo = TRUE}

setwd("C:/Users/acer17/Desktop/coursera/RR_Peer_Project1/RepData_PeerAssessment1")
```

The loaded row data "Activity"" contains 17568 observations of 3 variables. The cleaned data "subActivity" with removed "NA" values contains 15264 observations of 3 variables:
```{r readActivityData, echo = TRUE}
if (!file.exists("activity.csv")) { unzip(zipfile="./activity.zip") }
Activity <- read.csv("activity.csv")
subActivity <- subset(Activity, Activity$steps != "NA")
str(subActivity)
```

Load libraries:
```{r loadLibrariesData, echo=TRUE}
library(readr)
library(dplyr)
library(chron)
library(lattice)
```

## 1. What is mean total number of steps taken per day?

###Question: 
What is the mean total number of steps taken per day ?

###Answer: 
The mean of the total number of steps taken per day is: 10766.1886792453. The median of the total number of steps taken per day differs with 10765 a little bit. 


###Description
On target of this project is to investigate the influence from ignored missing values "NA", compared with filled missing values. For this part of the analysis, the missing values "NA"" where ignored.

Calculate the total numbers of steps taken each day:
```{r dataTotSteps, echo=TRUE}
sumSteps <- tapply(subActivity$steps,subActivity$date,sum) 
stepsPerDays <- as.data.frame(cbind(sumSteps, names(sumSteps)))
```

Make a histogram of the total number of steps taken each day:
```{r histTotSteps, echo=TRUE }
hist(as.numeric(as.character(stepsPerDays$sumSteps)),  xlab = "Total number of steps taken per day", ylab = "Number of days", main= "Most days (25) between 10'000 to 11'000 total steps taken per day", sub = "RR1_plot1: missing values NA ignored", col = "lightblue", freq = TRUE)
```


The distribution of the histogram above shows at most days there where a total of 10'000 to 11'000 steps taken per day.


Calculate and report mean and median of total number of steps taken per day:
```{r meanTotStep, echo=TRUE}
print(paste("Mean total number of steps taken per day:", mean(as.numeric(as.character(stepsPerDays$sumSteps)), na.rm = TRUE)))
print(paste("Median total number of steps taken per day:", median(as.numeric(as.character(stepsPerDays$sumSteps)), na.rm = TRUE)))
```


## 2. What is the average daily activity pattern?

###Question: 
Which 5-minute interval, on average across all days in the dataset, contains the maximum number of steps ?

###Answer: 
The 5-minute interval containing the highest number of steps is: 835



###Description
Makes a time serie plot of the mean number of steps taken per 5-minute interval, across all days in the dataset.

```{r dailyAveragePlot, echo=TRUE}
meanPerInterval <- tapply(subActivity$steps,subActivity$interval, mean) 
meanPerInterval <- as.data.frame(cbind(meanPerInterval, names(meanPerInterval)))
names(meanPerInterval) <- c("meanStepsInterval", "interval")
plot(as.numeric(as.character(meanPerInterval$interval)), as.numeric(as.character(meanPerInterval$meanStepsInterval)), xlab = "5-minute interval", ylab= "Mean of steps", col = "blue", type="l", main = "Intervals 835 with maximum number of steps", sub = "RR1_plot2")
```

The time serie plot above of the mean number of steps taken per interval, across all days in the dataset, shows increased activities from intervals 800 to 900, and nearly no activities from interval 0 to 525.

Calculate the 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps:
```{r maxAverageInterval, echo=TRUE}
stepsMax <- as.integer(as.character(meanPerInterval[meanPerInterval$meanStepsInterval==max(as.numeric(as.character(meanPerInterval$meanStepsInterval))), 2]))
print(paste("5-minute interval cotaining the maximum number of steps, calculated on average across all days:", stepsMax))
```


## 3. Imputing missing values

###Question: 
What is the impact of imputing missing data on the estimates of the total daily number of steps ?

###Answer: 
The distribution of the total daily number of steps tend now clearer to the mean value: 10766.1886792453. The histogram RR1_Plot3 has with 35 days more days with ~ 11000 steps taken per day then RR1_Plot1 with 25 days. 
And the mean value do not longer differ from the median value: 10766.1886792453. With ignored missing values, see Question 1, there was a difference between mean value: 10766.1886792453 and median value: 10765.



###Description
There are days/intervals where there are missing values (coded as NA). The presence of this missing values may introduce bias into some calculation or summaries of the data.From total 17568 observations there are 2304 observations with missing values.For this reason a new dataset "ActivityCorr" with the missing values (number of steps) where filled with the mean of steps for that interval. "ActivityCorr has the equal variables and numbers of observations like the origin "Activity" data. The variables included in this dataset are:

    steps       Number of steps taking in a 5-minute interval 
    
    date        The date on which the measurement was taken in YYYY-MM-DD format
    
    interval    Identifier for the 5-minute interval in which measurement was taken
    

Calculate and report the total number of missing values NA in the dataset, i.e. number of rows:
```{r nrMissingVal, echo=TRUE}
missingValue <- nrow(subset(Activity[is.na(Activity$steps) | is.na(Activity$interval) | is.na(Activity$date), ]))
print(paste("Total number of rows with missing values NA:", missingValue))
```
Replace missing values of steps with mean of number of steps for this interval, across all days. Create a new dataset that has equal data structure like the original dataset Activity with filled missing values:
```{r fillMissValues, echo = TRUE}
ActivityCorr <- merge(Activity, meanPerInterval, by="interval")
ActivityCorr[which(is.na(ActivityCorr$steps)) , 2] <- as.numeric(as.character(ActivityCorr[which(is.na(ActivityCorr$steps)) , 4])) 
ActivityCorr <- cbind(ActivityCorr, ActivityCorr$interval)
ActivityCorr <- subset(ActivityCorr[, c(2,3,5)])
colnames(ActivityCorr) <- c("steps", "date", "interval")
```

Calculate the total number of steps taken each day and plot a histogram of it:
```{r plotTotSteps, echo = TRUE}
sumStepsCorr <- tapply(ActivityCorr$steps,ActivityCorr$date,sum) 
stepsPerDaysCorr <- as.data.frame(cbind(sumStepsCorr, names(sumStepsCorr)))
hist(as.numeric(as.character(stepsPerDaysCorr$sumStepsCorr)),  xlab = "Total number of steps taken per day", ylab = "Number of days", main= "Most days (35) between 10'000 to 11'000 total steps taken per day", sub = " RR1_plot3: missing values filled by mean steps of interval", col = "grey", freq = TRUE)
```

Calculate mean and median of total number of steps taken per day, with missed values filled by mean of that interval:
```{r meanfilledValues, echo = TRUE}
print(paste("Mean of total number of steps taken per day with missed values filled by mean of interval:", mean(as.numeric(as.character(stepsPerDaysCorr$sumSteps)), na.rm = TRUE)))
print(paste("Median of total number of steps taken per day with missed values filled by median of interval:", median(as.numeric(as.character(stepsPerDaysCorr$sumSteps)), na.rm = TRUE)))
```

## 4. Are there differences in activity patterns between weekdays and weekends?

###Question: 
Are there differences in activity patterns between weekdays and weekends?

###Answer: 
Yes, there are differences between weekdays and weekend activity pattern. The activity pattern for weekdays shows fast increasing activities from  interval 500 to 600. At weekend there are very slow increasing activities from 500 to 800. So it seams the most persons don't like to stand up early if they dont have do work and their activitiy levels are more individual at weekend. 
From interval 900 to 2200 there is a higher activity level at weekend. People seams to use their freetime do move arround and less persons may be sitting in the office.
At weekday there is earlier in the evening less activity (~ 10 steps), at interval 2000 instead of interval 2200 at weekend. 



###Description
Average number of steps per 5-minute interval, taken across all weekday or weekends calculated. Dataset "ActivityCorr" with filled-in missing values used.


Create a factor variable with two levels "weekday" and "weekend":
```{r factorWeekday, echo=TRUE}
wdays <- factor(is.weekend(as.Date(ActivityCorr$date)), labels = c("weekday", "weekend"))
```
Make a panel plot of average number of steps per interval, taken across all weekday or weekends:
```{r plotTotWeekdays, echo = TRUE}
meanStepsWdayInt <- summarise(group_by(cbind(ActivityCorr, wdays), wdays, interval), meanSteps= mean(steps))
p <- xyplot(meanSteps ~ interval | wdays, data = meanStepsWdayInt, type="l", xlab = "Interval", ylab = "Number of steps", sub = "RR1_plot4", layout = c(1, 2))   
print(p)
```


